Uncovering underlying high-level musical content in the time domain \\ \small Leveraging self-supervised neural networks, inductive bias, and aural skills to learn deep audio embeddings with applications to boundary detection tasks

### 2023-08-05 abstract/abstract.tex
Updates from Overleaf

In collaboration with Epidemic Sound AB and the Music Technology Group (MTG) at Universitat Pompeu Fabra (UPF), we used self-supervised contrastive learning to uncover the underlying structure of Western tonal music by learning deep audio features for unsupervised music boundary detection. We applied deep convolutional neural networks with triplet loss function to identify abstract and semantic high-level musical elements without relying on their sonic qualities. This way, we replaced traditional acoustic features with deep audio embeddings, paving the way for sound-agnostic and content-sensitive music representation for boundary detection.

Our cognitively-based approach for learning embeddings focuses on using full-resolution data and preserving high-level musical information which unfolds in the time domain. A key component in our methodology is the use of triplet networks, which effectively understand and preserve the nuanced relationships within musical data. Drawing upon our domain expertise, we developed robust transformations to encode heuristic musical concepts that should remain constant. This novel approach aims to reconcile music and machine learning, enhancing machine listening models' efficacy through deep learning and triplet networks.

Preliminary results suggest that, while not outperforming state-of-the-art, our musically-informed technique has significant potential for boundary detection tasks and, most likely, so does for nearly all MIR downstream sound-agnostic and content-sensitive tasks constrained by data scarcity as it is possible to achieve competitive performance to traditional handcrafted signal processing methods by learning only from unlabeled audio files.

>>>>

In collaboration with Epidemic Sound AB and the Music Technology Group (MTG) at Universitat Pompeu Fabra (UPF), we used self-supervised contrastive learning to uncover the underlying structure of Western tonal music by learning deep audio features for music boundary detection. We applied deep convolutional neural networks with triplet loss function to identify abstract and semantic high-level musical elements without relying on their sonic qualities. This way, we replaced traditional acoustic features with deep audio embeddings, paving the way for sound-agnostic and content-sensitive music representation for downstream track segmentation tasks.

Our cognitively-based approach for learning embeddings focuses on using full-resolution data and preserving high-level musical information which unfolds in the time domain. A key component in our methodology is triplet networks, which effectively understand and preserve the nuanced relationships within musical data. Drawing upon our domain expertise, we developed robust transformations to encode heuristic musical concepts that should remain constant. This novel approach aims to reconcile music and machine learning, enhancing machine listening models' efficacy.

Preliminary results suggest that, while not outperforming state-of-the-art, our musically-informed technique has significant potential for boundary detection tasks. Most likely, so does for nearly all downstream sound-agnostic and content-sensitive tasks constrained by data scarcity, as it is possible to achieve competitive performance to traditional handcrafted signal processing methods by learning only from unlabeled audio files.

### 2023-08-05 acknowledgements/acknowledgements.tex
Updates from Overleaf

\begin{itemize}
\item Carl Thomé and Carlos Lordelo, whose expertise in MIR, ML, and DSP has been pivotal to the success of my thesis. Their guidance and encouragement have enabled me to overcome the challenges encountered during this journey. Their commitment to mentorship is truly inspiring, and their influence has significantly shaped my academic path. 
Working under their mentorship has been both a privilege and an honor.

\vspace*{3mm}
\item Cody Hesse and Sebastian Löf. Their generosity in sharing their time, knowledge, and insights has critically enhanced the quality of this research work.

\vspace*{3mm}
\item The Music Technology Group (MTG) and Universitat Pompeu Fabra (UPF). My profound thanks go to Xavier Serra for his faith in accepting a non-traditional profile into the master's program. I am equally grateful to Pedro Ramoneda and Genís Plaja for their consistent support throughout my journey.

@@

I sincerely thank you all for your integral roles in completing my thesis. I deeply value the knowledge and experience gained under your guidance and will cherish it always.

>>>>

\begin{itemize}
\item Carl Thomé and Carlos Lordelo, supervisors and fellow MIR researchers. Their expertise has been pivotal to the success of this thesis. Their guidance, supervision, and encouragement have enabled me to overcome the challenges encountered during this journey. Their commitment to mentorship is genuinely inspiring, and their influence has significantly shaped my academic path.

Working with them has been an absolute privilege.

\vspace*{3mm}
\item Cody Hesse and Sebastian Löf, fellow interns. Their generosity in sharing their time, knowledge, and insights has critically enhanced the quality of this research work.

\vspace*{3mm}
\item The Music Technology Group (MTG) and Universitat Pompeu Fabra (UPF). My profound thanks go to Xavier Serra for his faith in accepting a non-traditional profile into the master's program.

@@

I sincerely thank you for your integral roles in completing this thesis. I deeply value the knowledge and experience gained under their guidance and will cherish it always.

### 2023-08-05 discussion/discussion.tex
Updates from Overleaf

While our musically-informed technique does not yet outperform the existing state-of-the-art baselines \cite{deepfeaturesegment, SalamonDeepSegmentation}, it exhibits significant potential in boundary detection tasks, particularly when expanding the training set. The improvement we've observed between datasets is noteworthy, and when compared with two of the acoustic features, our results are highly competitive. We've also managed to circumvent the typical issues associated with dataset enlargements, such as the need for extensive supervision or human annotation, which gives our method an edge in terms of practicality and scalability, effectively turning what is usually seen as an expensive hurdle into a much more manageable task.

@@

This represents a 'thorn in our side' that we intend to address in future research.

We argue whether factors such as the size of the representation layer and the size of the model \cite{verydeep} might be insufficient. Furthermore, we posit that with additional time to iterate repeatedly over the dataset, the loss function could decrease further with minimal risk of overfitting. \ref{fig:lossf}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{figures/images/Mean group aggregation.png}
    \caption[Loss function performance plot]{Mean group aggregation of the decrease of the loss function over trainer/global steps. Training loss against validation loss.}
    \label{fig:lossf}
\end{figure}

>>>>

While our musically-informed technique does not yet outperform the existing state-of-the-art baselines, it exhibits significant potential in boundary detection tasks, particularly when expanding the training set. The improvement we've observed between datasets is noteworthy, and when compared with two of the acoustic features, our results are highly competitive.

We have also managed to circumvent the typical issues associated with dataset enlargements, such as the need for extensive supervision or human annotation, which gives our method an edge in terms of practicality and scalability, effectively turning what is usually seen as an expensive hurdle into a much more manageable task.

@@

We argue whether factors such as the size of the representation layer and the size of the model \cite{verydeep} might be insufficient. Furthermore, we posit that the loss function could decrease further with additional time to iterate repeatedly over the dataset.

@@

### 2023-08-05 introduction/Subsections/Introduction.tex
Updates from Overleaf

Music \footnote{Throughout this thesis, "music" refers to the Western tradition, defined by the conventions, practices, and aesthetics that primarily evolved in Europe and North America. This term's usage does not negate the diversity and richness of other musical traditions worldwide. Instead, it specifies this study's concentration on a particular cultural context.}, a cornerstone of human culture and history, has developed hand in hand with our societies since ancient times. However, its study presents unique challenges due to its subjectivity, fluctuating ground truth, and the vast array of styles and cultural contexts it encompasses.

@@

Beyond applications in music structure analysis, the evolution of neural networks has empowered the creation of latent representation encapsulating crucial musical traits, technically known as representation (or feature) learning. This evolution has simplified the computational process for various task-specific elements, including but not limited to:

@@

Given the proven effectiveness of deep audio embeddings in existing research and their potential for transfer learning, they hold considerable promise for MIR downstream tasks, including boundary detection. However, it is yet to be determined whether a general-purpose audio representation can successfully emulate human hearing \cite{Turian2022HEAR:Representations}, even though some techniques have demonstrated generalizability across many music-understanding tasks \cite{Li2023MERT:Training, Kim2020OneStrategies}.

Yet, despite these strides, deep learning's role in music understanding remains relatively nascent, with scarce work on deep music representations, a dearth of large-scale datasets, and a lack of a universal community-driven benchmark \cite{Turian2022HEAR:Representations, Yuan2023MARBLE:Evaluation}.

>>>>

Music \footnote{Throughout this thesis, "music" refers to the Western tradition, defined by the conventions, practices, and aesthetics that primarily evolved in Europe and North America. This term's usage does not negate the diversity and richness of other musical traditions worldwide. Instead, it specifies this study's concentration on a particular cultural context.}, a cornerstone of human culture, has developed hand in hand with our societies since ancient times. However, its study presents unique challenges due to its subjectivity, fluctuating ground truth, and the vast array of styles and cultural contexts.

@@

Beyond applications in music structure analysis, the evolution of neural networks has empowered the creation of latent representations \ref{fig:manifold} encapsulating crucial musical traits, technically known as representation ---or feature--- learning. This evolution has simplified the computational process for various task-specific elements, including but not limited to:

@@

Unsupervised settings to learn those embeddings confront the persistent challenge of acquiring labeled data, an endeavor often time-consuming and expensive. By capitalizing on the vast volumes of available unlabeled music data, we bypass the need for laborious hand annotation, offering a pathway to potentially more comprehensive generalization.

It is yet to be determined whether a general-purpose audio representation can successfully emulate human hearing \cite{Turian2022HEAR:Representations}, even though some techniques have demonstrated generalizability across many music-understanding tasks \cite{Li2023MERT:Training, Kim2020OneStrategies}. Deep learning's role in music understanding remains relatively nascent, with scarce work on deep music representations, a dearth of large-scale datasets, and a lack of a universal community-driven benchmark \cite{Turian2022HEAR:Representations, Yuan2023MARBLE:Evaluation}.

All in all, given the proven effectiveness of deep audio embeddings in existing research and their potential for transfer learning, they hold considerable promise; therefore, we will leverage deep convolutional neural net-
works to replace traditional
acoustic features with sound-agnostic
and content-sensitive embeddings with applications to boundary detection and tracking segmentation tasks

### 2023-08-05 introduction/Subsections/Objectives.tex
Updates from Overleaf

Our research aims to develop and learn a unified numerical understanding (or embedding \footnote{A learned representation or embedding is a numerical output - usually a fixed-size vector - produced by a machine-learning model. Good representations should be versatile across various tasks and require limited supervision \cite{Turian2022HEAR:Representations}.}) for a piece of music that integrates the high-level relationships between musical elements as they unfold over time.

The research employs MIR computational methodologies and musicological perspectives to explore musical compositions across various styles and genres.

@@

>>>>

Our research aims to develop and learn a unified numerical understanding ---or embedding \footnote{A learned representation or embedding is a numerical output - usually a fixed-size vector - produced by a machine-learning model. Good non-supervised representations should be versatile across various tasks and require limited supervision, therefore appears as an attractive methodology to be employed \cite{Turian2022HEAR:Representations}.}--- for a piece of music that integrates the high-level relationships between elements as they unfold over time.

@@

\section{Contributions}

Our study is centered around the following topics of investigation:

\begin{itemize}
    \item We demonstrate that the size of the audio file dataset used for learning embeddings directly influences the results of music segmentation. Larger datasets lead to improved results.
    \item We show that achieving competitive performance with traditional handcrafted signal processing methods is possible by learning solely from unlabeled audio files.
    \item While our musically-informed technique does not currently surpass existing state-of-the-art baselines, it shows significant promise in boundary detection tasks, especially when the training set is expanded.
\end{itemize}

### 2023-08-05 methods/subsections/deep learning.tex
Updates from Overleaf

DNNs, computational models somewhat rooted in biological neural systems, utilize layers of artificial neurons for tasks such as pattern recognition, classification, and regression. They evolve by adjusting connections between neurons during learning, a concept initiated in the mid-20th century but modernized with Frank Rosenblatt's single-layer perceptron in 1957, capable of learning linearly separable patterns \cite{perceptron}.

@@

Another crucial component in the operation of DNNs is the activation function. This function is applied at every layer, transforming the linear combination of the input with the layer weights into an output that is passed onto the next layer. It introduces non-linearity into the network, enabling it to learn complex patterns and relationships from the data.

>>>>

These computational models somewhat rooted in biological neural systems, utilize layers of artificial neurons for tasks such as pattern recognition, classification, and regression. They evolve by adjusting connections between neurons during learning, a concept initiated in the mid-20th century but modernized with Frank Rosenblatt's single-layer perceptron in 1957, capable of learning linearly separable patterns \cite{perceptron}.

@@

Another crucial component in the operation of DNNs is the activation function. This function is applied at every layer, transforming the linear combination of the input with the layer weights into an output that is passed onto the next layer. It introduces non-linearity into the network, enabling it to learn complex patterns and relationships.

### 2023-08-05 preface/preface.tex
Updates from Overleaf

When the opportunity to work at Epidemic Sound presented itself, it was as if the universe had conspired to bring my interests together, offering me the chance to explore the fascinating field of Music Information Retrieval (MIR) from an industrial perspective.

@@

This work is intended for any curious person, regardless of their academic background. I hope it will inspire others to explore the fascinating intersection of these two disciplines.

>>>>

When the opportunity to work at Epidemic Sound presented itself, it was as if the universe had conspired to bring my interests together, allowing me to explore the fascinating Music Information Retrieval (MIR) field from a top-notch industrial perspective.

@@

This work is intended for any curious person, regardless of their background. I hope it will inspire others to explore the fascinating intersection of these two disciplines.
